Towards a Verified Tableau Prover for Multi-Level-Syllogistics

### 2022-09-16 paper/paper.tex
Continue

\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
We combine these functions in order to define \lstinline!vars! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables, Mention vars polymorphic}
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}
In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: linear rules don't introduce subterms, \\ \lstinline!subterms b $\approx$ subterms (last b)!}
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

\subsection{Linear Expansion Rules}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
Furthermore, linear rules are restricted such that no new subterm may be created by their application:
for example, the rule
\begin{center}
\lstinline!(s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&!
\end{center}
may only be used under the condition
\begin{center}
  \lstinline!t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b)!.
\end{center}
\todo[inline]{Explain this restriction in more detail.}
This ensures that repeated application of linear rules eventually results in a \textit{linearly saturated} branch, i.e.\ a branch where no application of linear rules would produce new formulae.
In Isabelle, we express this as follows.
\begin{lstlisting}
definition lin_sat b $\equiv$
  $\forall$b'. &\lexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}

\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{p{0.21\textwidth}cp{0.18\textwidth} p{0.01\textwidth} p{0.21\textwidth}cp{0.18\textwidth}}
    \multicolumn{3}{c}{Propositional Rules} && \multicolumn{3}{c}{Rules for \lstinline!$\unionS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! &&
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\

@@

    \multicolumn{3}{c}{Rules for \lstinline!Single!} && \multicolumn{3}{c}{Rules for \lstinline!$\eqS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, ($\fmNegSymbol$) a! & $\Longrightarrow$ & \lstinline!($\fmNegSymbol$) a[$\ttw$/$\ton$]! \\

    \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, ($\fmNegSymbol$) a! & $\Longrightarrow$ & \lstinline!($\fmNegSymbol$) a[$\ton$/$\ttw$]! \\

    \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! &&
    \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\inS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
\todo[inline]{Reference figure and mention missing rule + example}

\subsection{Branching Rules\label{sec:branching}}
The number of branches increases when applying branching rules so it is prudent to only apply them to open, linearly saturated branches.
To further prevent unnecessary branching, each rule has a \textit{subsumption} condition. 
Note that even branching rules are restricted such that they never create new subterms, with exception of the last rule which adds a fresh variable.
\todo[inline]{Reference figure}

\begin{lstlisting}
definition sat b $\equiv$ lin_sat b $\land$ ($\nexists$bs'. &\fexpands{bs'}{b}&)
\end{lstlisting}

\begin{lstlisting}
definition wits :: 'a branch $\Rightarrow$ 'a set where
  wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.28\textwidth} >{\centering\arraybackslash}p{0.3\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & Test & \lstinline!p $\in$ set b $\lor$ @\fmNeg{p}@ $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!@\fmNeg{p}@! & \lpipe & \lstinline!p! \\
    \end{tabular}
         & Test & \lstinline!@\fmNeg{p}@ $\in$ set b $\lor$ p $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ton$! & \lpipe & \lstinline!s $\notinS$ $\ton$! \\
    \end{tabular}
         & Test &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ton$) $\in$ set b!
         \end{tabular}
         \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         & Test &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
        \end{tabular}
        \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         & Test &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
         \end{tabular} \\
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!Var x $\inS$ $\ton$! \lstinline!Var x $\notinS$ $\ttw$! & \multirow{2}{*}{\rule[-0.4ex]{0.6pt}{4.8ex}} & \lstinline!Var x $\notinS$ $\ton$! \lstinline!Var x $\inS$ $\ttw$! \\
    \end{tabular}
         & Test & Test \\
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands}}
\end{table*}

@@

\section{Upper Bounding the Cardinality of a Branch\label{sec:bound}}
Soundness and completeness of the calculus are the building blocks with which we establish partial correctness of the algorithm in \autoref{fig:alg}.
Consequently, we still need to give a termination proof to ascertain total correctness.
The core of our argument is that the number of distinct formulae that can be derived with the expansion rules is finite.
In particular, we claim that the following upper bound holds.
\begin{lstlisting}
  &\phantom{$\leq$}& card (set b)
  &\textover{$=$}{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$} +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& 16 * (card (subterms (last b)))$^4$
\end{lstlisting}
We treat both inequalities separately, starting with the first one whose justification hinges on an invariant: 
every new composite subformula that is introduced by an expansion rule is the negation of a subformula that already occurs in \lstinline!last b!.
Written as a formula, we have
\begin{lstlisting}
  &\phantom{$\subseteq$}& {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\subseteq$ subfms (last b) $\cup$ $\fmNegSymbol$ ` subfms (last b).
\end{lstlisting}
Confirming that this is an invariant is an easy exercise of going through each of the propositional rules in Table TODO and the first two branching rules in Table TODO.
Thus, we have
\begin{lstlisting}
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\leq$ card (subfms (last b)) +
  &\phantom{$\leq$}& card ($\fmNegSymbol$ ` subfms (last b))
  $\leq$ 2 * card (subfms (last b))
\end{lstlisting}
which is the first inequality we asserted.

>>>>

\begin{lstlisting}
fun is_literal :: 'a fm $\Rightarrow$ bool where
  is_literal (&\fmAtom& _) = True
| is_literal (&\fmNeg{(\fmAtom{} \_)}&) = True
| is_literal _ = False
\end{lstlisting}
We will often drop the atom constructor \lstinline!&\fmAtom{}&! to reduce clutter.
Additionally, we use the abbreviations \lstinline!s $\notinS$ t! and \lstinline!s $\neqS$ t! to denote \lstinline!&\fmNeg{\fmAtom{} (s $\inS$ t)}&! and \lstinline!&\fmNeg{\fmAtom{} (s $\eqS$ t)}&!, respectively.
Note that Isabelle's datatype package automatically defines a function \lstinline!atoms :: 'a fm $\Rightarrow$ 'a set! that allows us to obtain all set atoms in a formula.
We combine these functions in order to define \lstinline!vars! that extracts all the variable names occuring in a set formula.
\todo[inline]{Motivation: one rule introduces new variables, Mention vars polymorphic}
\begin{lstlisting}
definition vars :: 'a pset_fm $\Rightarrow$ 'a set where
  vars $\phi$ $\equiv$ $\bigcup$(vars ` atoms $\phi$)
\end{lstlisting}

In addition to the variable names that appear in a formula, we will later also need to obtain those set terms that occur as subterms in a formula. 
We first define a function that collects the subterms of a set term.
\todo[inline]{Motivation: linear rules don't introduce subterms, \\ \lstinline!subterms b $\approx$ subterms (last b)!}
\begin{lstlisting}
fun subterms :: 'a pset_term
             $\Rightarrow$ 'a pset_term set where
  subterms $\emptyset$ = {$\emptyset$}
| subterms (Var x) = {Var x}
| subterms (s $\unionS$ t) =
    {s $\unionS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\interS$ t) =
    {s $\interS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (s $\diffS$ t) =
    {s $\diffS$ t} $\cup$ subterms s $\cup$ subterms t
| subterms (Single t) =
    {Single t} $\cup$ subterms t
\end{lstlisting}
The subterms of an atom are then the subterms of those terms that make up the atom.
\begin{lstlisting}
fun subterms :: 'a pset_atom
             $\Rightarrow$ 'a pset_term set where
  subterms (s $\inS$ t) = subterms s $\cup$ subterms t
| subterms (s $\eqS$ t) = subterms s $\cup$ subterms t
\end{lstlisting}
Ultimately, we lift this to formulae by using \lstinline!atoms! again.
\begin{lstlisting}
definition subterms ::
    'a pset_fm $\Rightarrow$ 'a pset_term set where
  subterms $\phi$ $\equiv$ $\bigcup$(subterms ` atoms $\phi$)
\end{lstlisting}
The last function that we define computes the subformulae of a formula.
\begin{lstlisting}
fun subfms :: 'a fm $\Rightarrow$ 'a fm set where
  subfms (&\fmAtom& a) = {&\fmAtom& a}
| subfms (&\fmAnd{p}{q}&) =
    {&\fmAnd{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmOr{p}{q}&) =
    {&\fmOr{p}{q}&} $\cup$ subfms p $\cup$ subfms q
| subfms (&\fmNeg{q}&) = {&\fmNeg{q}&} $\cup$ subfms q
\end{lstlisting}

@@

\subsection{Linear Expansion Rules}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \begin{tabular}{p{0.21\textwidth}cp{0.18\textwidth} p{0.01\textwidth} p{0.21\textwidth}cp{0.18\textwidth}}
    \toprule
    \multicolumn{3}{c}{Propositional Rules} && \multicolumn{3}{c}{Rules for \lstinline!$\unionS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    \lstinline!@\fmAnd{p}{q}@! & $\Longrightarrow$ & \lstinline!p, q! &&
    \lstinline!s $\notinS$ $\ton$ $\unionS$ $\ttw$! & $\Longrightarrow$ & \lstinline!s $\notinS$ $\ton$, s $\notinS$ $\ttw$! \\

@@

    \multicolumn{3}{c}{Rules for \lstinline!Single!} && \multicolumn{3}{c}{Rules for \lstinline!$\eqS$!} \\
    \cmidrule{1-3}\cmidrule{5-7}
    & $\Longrightarrow$ & \lstinline!s $\inS$ Single s! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ttw$/$\ton$}! \\

    \lstinline!s $\inS$ Single t! & $\Longrightarrow$ & \lstinline!s $\eqS$ t! &&
    \lstinline!$\ton$ $\eqS$ $\ttw$, l! & $\Longrightarrow$ & \lstinline!l{$\ton$/$\ttw$}! \\

    \lstinline!s $\notinS$ Single t! & $\Longrightarrow$ & \lstinline!s $\neqS$ t! &&
    \lstinline!s$_\text{1}$ $\inS$ t, s$_\text{2}$ $\inS$ t! & $\Longrightarrow$ & \lstinline!s$_\text{1}$ $\neqS$ s$_\text{2}$! \\
    \bottomrule
  \end{tabular}
  \bigskip
  \caption{Linear expansion rules.\label{tab:lexpands}}
\end{table*}
The calculus considers two kinds of branch expansion rules: \textit{linear} and \textit{branching} rules.
As the name suggests, branching rules lead to the creation of new branches in the tableau while linear rules only extend a branch \lstinline!b! with new formulae \lstinline[breaklines=true]!b' = [$\psi_1$,$\ldots$,$\psi_n$]!, which we denote by \lstinline!&\lexpands{b'}{b}&!.
\autoref{tab:lexpands} shows all linear expansion rules.
Note that, in the first two rules for \lstinline!$\eqS$!, \lstinline!l! is a literal occuring in the branch.
Furthermore, the term-for-term substitution \lstinline!l{s/t}! is restricted to the top-level set terms of the literal;
for example, given the literal
\begin{lstlisting}
  l = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&
\end{lstlisting}
we have
\begin{lstlisting}
  &\phantom{=}& (&\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ s $\unionS$ u)}&){t/s $\unionS$ u}
  = &\fmNeg{((s $\unionS$ u) $\diffS$ s $\eqS$ t)}&.
\end{lstlisting}
The perhaps more crucial restriction of the linear rules is that that no new subterm may be created by their application;
for instance, the second rule for \lstinline!$\unionS$!
\begin{lstlisting}
  s $\inS$ t$_\text{1}$ $\Longrightarrow$ s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$,
\end{lstlisting}
which formally represents 
\begin{lstlisting}
  (s $\inS$ t$_\text{1}$) $\in$ set b $\Longrightarrow$ &\lexpands{[s $\inS$ t$_\text{1}$ $\unionS$ t$_\text{2}$]}{b}&,
\end{lstlisting}
may only be used under the condition
\begin{lstlisting}
  t$_\text{1}$ $\unionS$ t$_\text{2}$ $\in$ subterms (last b).
\end{lstlisting}
The purpose of this restriction is to prevent unbounded expansion of the branch.
In fact, we give an explicit upper bound for the number of formulae in a branch in \autoref{sec:bound}.

Due to boundedness, repeated application of linear rules eventually results in a \textit{linearly saturated} branch, i.e.\ a branch where no application of linear rules would produce new formulae.
\begin{lstlisting}
definition lin_sat b $\equiv$
  $\forall$b'. &\lexpands{b'}{b}& $\longrightarrow$ set b' $\subseteq$ set b
\end{lstlisting}

Finally, we remark that the original paper~\cite{new_fast_tableau} is missing the last propositional rules dealing with double negation.
This rule is required for completeness, though, considering that the branch
\begin{lstlisting}
  [&\fmNeg{\fmNeg{\fmNeg{p}}}&, p, &\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&]
\end{lstlisting}
is saturated --- neither linear nor branching rules apply --- and open but there is obviously no model for the initial formula \lstinline!&\fmAnd{\fmNeg{\fmNeg{\fmNeg{p}}}}{p}&!.

\subsection{Branching Rules\label{sec:branching}}
\begin{table*}
  \lstset{
    escapeinside={@}{@}
  }
  \newcommand{\ton}{\text{t}_\text{1}}
  \newcommand{\ttw}{\text{t}_\text{2}}
  \newcommand{\lpipe}{\rule[-0.4ex]{0.6pt}{2.3ex}}
  \begin{tabular}{>{\centering\arraybackslash}p{0.28\textwidth} >{\centering\arraybackslash}p{0.34\textwidth} >{\centering\arraybackslash}p{0.3\textwidth}}
    \toprule
    Rule & Precondition & Subsumption condition \\
    \midrule
    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!p! & \lpipe & \lstinline!@\fmNeg{p}@! \\
    \end{tabular}
         & \lstinline!@\fmOr{p}{q}@ $\in$ set b! & \lstinline!p $\in$ set b $\lor$ @\fmNeg{p}@ $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!@\fmNeg{p}@! & \lpipe & \lstinline!p! \\
    \end{tabular}
         & \lstinline!@\fmNeg{(\fmAnd{p}{q})}@ $\in$ set b!
         & \lstinline!@\fmNeg{p}@ $\in$ set b $\lor$ p $\in$ set b! \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ton$! & \lpipe & \lstinline!s $\notinS$ $\ton$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$ $\unionS$ $\ttw$) $\in$ set  b! \\
           \lstinline!$\ton$ $\unionS$ $\ttw$ $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ton$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ton$) $\in$ set b!
         \end{tabular}
         \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$ $\interS$ $\ttw$) $\in$ set  b! \\
           \lstinline!($\ton$ $\interS$ $\ttw$) $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
        \end{tabular}
        \\

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!s $\inS$ $\ttw$! & \lpipe & \lstinline!s $\notinS$ $\ttw$! \\
    \end{tabular}
         &
         \begin{tabular}{c}
           \lstinline!(s $\inS$ $\ton$ $\diffS$ $\ttw$) $\in$ set  b! \\
           \lstinline!($\ton$ $\diffS$ $\ttw$) $\in$ subterms (last b)! \\
         \end{tabular}
         &
         \begin{tabular}{@{}rl@{}}
                  & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
           $\lor$ & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
         \end{tabular} \\
         \addlinespace[0.6em]

    \begin{tabular}{@{}>{\raggedleft\arraybackslash}p{0.11\textwidth}c>{\raggedright\arraybackslash}p{0.11\textwidth}@{}}
      \\
      \midrule
      \lstinline!Var x $\inS$ $\ton$! \lstinline!Var x $\notinS$ $\ttw$! & \multirow{2}{*}{\rule[-0.4ex]{0.6pt}{4.8ex}} & \lstinline!Var x $\notinS$ $\ton$! \lstinline!Var x $\inS$ $\ttw$! \\
    \end{tabular}
         & 
         \begin{tabular}{c}
           \lstinline!($\ton$ $\neqS$ $\ttw$) $\in$ set b! \\
           \lstinline!$\ton$ $\in$ subterms (last b)! \\
           \lstinline!$\ttw$ $\in$ subterms (last b)! \\
           \lstinline!x $\notin$ vars b! \\
         \end{tabular}
         &
         \begin{tabular}{r@{\hspace{0.5em}}l}

           \lstinline!$\exists$s.! & \lstinline!(s $\inS$ $\ton$) $\in$ set b $\land$! \\
                                   & \lstinline!(s $\notinS$ $\ttw$) $\in$ set b! \\
           \multicolumn{2}{c}{$\lor$} \\
           \lstinline!$\exists$s.! & \lstinline!(s $\notinS$ $\ton$) $\in$ set b $\land$! \\
                                   & \lstinline!(s $\inS$ $\ttw$) $\in$ set b! \\
         \end{tabular}
         \\
         \bottomrule
  \end{tabular}
  \bigskip
  \caption{Branching expansion rules.\label{tab:bexpands}}
\end{table*}
After running out of linear rules apply, only the branching rules shown in \autoref{tab:bexpands} remain.
A rule is applicable if its \textit{precondition} is met and, to prevent unnecessary branching, it is not subsumed as indicated by the \textit{subsumption condition}.
These rules create multiple branches in the tableau so we represent the different possibilities \lstinline!bs'! to expand a branch \lstinline!b! as a set and write \lstinline!&\fexpands{bs'}{b}&!. 
Accordingly, we get a new branch \lstinline!b' @ b! in the tableau for each \lstinline!b' $\in$ set bs'!.

A linearly saturated branch where no further branching is possible is called a \textit{saturated} branch. 
\begin{lstlisting}
definition sat b $\equiv$ lin_sat b $\land$ ($\nexists$bs'. &\fexpands{bs'}{b}&)
\end{lstlisting}

Note that even branching rules are defined such that they never create new subterms with exception of the last rule, which adds a new variable to the branch.
These variables serve to manifest an inequality so we call them \textit{witnesses}.
\begin{lstlisting}
definition wits :: 'a branch $\Rightarrow$ 'a set where
  wits b $\equiv$ vars b - vars (last b)
\end{lstlisting}

@@

\section{Upper Bounding the Cardinality of a Branch\label{sec:bound}}
Soundness and completeness of the calculus are the building blocks with which we establish partial correctness of the algorithm in \autoref{fig:alg}.
Consequently, we still need to give a termination proof to ascertain total correctness.
The core of our argument is that the number of distinct formulae that can be derived with the expansion rules is finite.
In particular, we claim that the following upper bound holds.
\begin{lstlisting}
  &\phantom{$\leq$}& card (set b)
  &\textover{$=$}{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$} +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. is_literal $\psi$}
  $\leq$ 2 * card (subfms (last b)) +
  &\phantom{$\leq$}& 16 * (card (subterms (last b)))$^4$
\end{lstlisting}
We treat both inequalities separately, starting with the first one whose justification hinges on an invariant: 
every new composite subformula that is introduced by an expansion rule is the negation of a subformula that already occurs in \lstinline!last b!.
Written as a formula, we have
\begin{lstlisting}
  &\phantom{$\subseteq$}& {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\subseteq$ subfms (last b) $\cup$ $\fmNegSymbol$ ` subfms (last b).
\end{lstlisting}
Confirming that this is an invariant is an easy exercise of going through each of the propositional rules in \autoref{tab:lexpands} and the first two branching rules in \autoref{tab:bexpands}.
Thus, we have
\begin{lstlisting}
  &\phantom{$\leq$}& card {$\psi$ $\in$ set b. $\neg$ is_literal $\psi$}
  $\leq$ card (subfms (last b)) +
  &\phantom{$\leq$}& card ($\fmNegSymbol$ ` subfms (last b))
  $\leq$ 2 * card (subfms (last b))
\end{lstlisting}
which is the first inequality we asserted.
